---
title: "Germany’s Hidden Burden: Healthcare-Associated Infections Compared to Europe"
author: "Nhu Nguyen"
date: 2025-09-20
categories: [health, analytics, assignment]
bibliography: references.bib
listing:
  contents:
    - posts/welcome/*.qmd
  sort: "date desc"
format:
  html:
    toc: true
    toc-depth: 3       
    toc-location: left  
    number-sections: false

---

# Problem description

Healthcare-associated infections (HAIs) are illnesses acquired during hospital care. They increase treatment costs, prolong hospital stays, and may lead to disability or death.

Although Germany’s hospital prevalence of HAIs is lower than the EU/EEA average, the population-level burden is higher. This apparent paradox arises from Germany’s *high hospital admission rates* and *longer average lengths of stay*.

This blog post analyses data from the ECDC Point Prevalence Survey (2011 - 2012) to explain Germany’s elevated burden and identify which infection types matter most. The five HAI types considered account for **≈79%** of HAIs in the PPS frame:\
- Clostridioides difficile infection (CDI)\
- Hospital-acquired pneumonia (HAP)\
- Bloodstream infection (BSI)\
- Urinary tract infection (UTI)\
- Surgical site infection (SSI)

# Analysis

## Data source

This analysis draws on the *European Centre for Disease Prevention and Control (ECDC) Point Prevalence Survey, 2011–2012*. 
Burden estimates were calculated using the *BCoDE methodology* (Burden of Communicable Diseases in Europe), implemented in the *BHAI R package* [@willrich2019bhai].


## Germany vs EU/EEA

@tbl-burden shows annual cases, deaths, and disability-adjusted life years (DALYs) per 100,000 population. Germany exceeds the EU/EEA average in all measures.

```{r, message=FALSE, warning=FALSE}
#| label: tbl-burden
#| tbl-cap: "Estimated HAI burden per 100,000 population, Germany vs EU/EEA (2011–2012)"
#| message: false
#| warning: false
#| code-fold: true

burden <- data.frame(
  Metric   = c("Cases (per 100k)", "Deaths (per 100k)", "DALYs (per 100k)"),
  Germany  = c(592, 20.1, 308),
  EU_EEA   = c(468, 15.6, 290)
)

knitr::kable(
  burden,
  caption = NULL,
  align = c("l","r","r"),
  booktabs = TRUE
)


```

@fig-types displays DALYs for major HAI types in Germany. Hospital-acquired pneumonia (HAP) and bloodstream infections (BSI) dominate mortality, while urinary tract infections (UTI) and surgical site infections (SSI) contribute to disability.

```{r}
#| label: fig-types
#| fig-cap: "Germany’s HAI burden by infection type (DALYs, German PPS 2011–2012)"
#| fig-width: 7
#| fig-height: 4.5
#| message: false
#| warning: false
#| code-fold: true

library(ggplot2)

types_totals <- data.frame(
  Infection = c("HAP","BSI","UTI","SSI","CDI"),
  DALYs = c(69508, 58350, 66701, 28842, 20890)
)

ggplot(types_totals, aes(x = Infection, y = DALYs)) +
  geom_col(fill = "#274060") +
  geom_text(aes(label = scales::comma(DALYs)),
            vjust = -0.35, size = 3.5) +
  labs(title = "DALYs by infection type in Germany (absolute)",
       x = "Infection type",
       y = "DALYs (years)",
       caption = "Source: ECDC PPS 2011–2012; BHAI estimates (German PPS sample).") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", color = "#005c97"),
    axis.text.x = element_text(angle = 10, hjust = 1),
    panel.grid.minor = element_blank()
  )

```

@tbl-type-rates breaks this down by infection type.  
- Germany’s case rate for **UTI** is ~1.5× higher than the EU/EEA average (265 vs 175 per 100 k).  
- **CDI** and **BSI** show notably higher fatality and DALY rates.  
- **HAP** rates are slightly lower in Germany, yet the absolute burden remains large due to long stays and recurrent pneumonia among older inpatients.  

```{r}
#| label: tbl-type-rates
#| tbl-cap: "HAI burden per 100,000 by infection type: Germany (PPS) vs EU/EEA (ECDC PPS), 2011–2012"
#| message: false
#| warning: false
#| code-fold: true

types_rates <- data.frame(
  Infection = c("HAP","UTI","BSI","SSI","CDI","ALL"),
  Germany_cases = c(132.0, 265.1, 33.4, 115.4, 44.6, 592.1),
  EU_EEA_cases  = c(143.7, 174.7, 22.2, 111.3, 16.0, 467.9),
  Germany_deaths = c(4.9, 4.5, 4.8, 2.9, 2.4, 20.1),
  EU_EEA_deaths  = c(5.3, 3.0, 3.3, 2.6, 0.9, 15.3),
  Germany_DALYs  = c(86.1, 82.6, 72.2, 35.7, 25.9, 308.2),
  EU_EEA_DALYs   = c(109.8, 57.1, 76.2, 35.1, 10.0, 290.0)
)

knitr::kable(
  types_rates,
  caption = NULL,
  align = c("l","r","r","r","r","r","r"),
  booktabs = TRUE,
  col.names = c("Infection","Cases DE","Cases EU","Deaths DE","Deaths EU","DALYs DE","DALYs EU")
)

```

Together, these tables demonstrate that Germany’s excess burden stems not from one pathogen but from **system-wide exposure and patient throughput**: more admissions and longer hospital stays amplify even modest infection risks.

## Findings

-   Germany had ≈478,000 cases, 16,000 deaths, and 249,000 DALYs from HAIs in 2011–2012

-   HAP and BSI caused more than half the total burden despite fewer cases.

-   UTI was the most common infection but had lower fatality.

-   SSI contributed disproportionately to disability.

-   HAIs imposed a greater health burden than influenza, salmonella, hepatitis B, and measles combined

These findings link back to the problem statement: Germany’s system-level hospital practices amplify the burden despite lower prevalence.

# Conclusion

Germany’s elevated HAI burden is primarily system-driven rather than patient-driven.  
Lower per-patient prevalence is outweighed by:
1. High hospital admission rates, and  
2. Prolonged inpatient durations.

The analysis shows that **HAP** and **BSI** drive mortality, while **UTI** and **SSI** dominate morbidity and disability.  
This pattern mirrors findings by Cassini et al. (2016) [@cassini2016] and Zacher et al. (2019) [@zacher2019], confirming that the national healthcare structure, rather than clinical hygiene alone that underlies the disproportionate population burden.

**Key takeaway:** To reduce the national burden of HAIs, Germany must pair infection control measures with *systemic reforms* that shorten hospital stays and optimize admission practices.
